Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-12-23 ms.tex
Expand Ref. abbreviation at beginning of sentence
Break up some equations across multiple lines
Fix some typos found in proof

A more natural way is to plug the signal model from Eq.~(\ref{eq:signal-model}) evaluated at the \ac{ML} parameter estimates into the single-detector likelihood in Eq.~(\ref{eq:gaussian-likelihood-spa}):
%
\begin{widetext}
\begin{equation}
    p\left(\hat{\bm\theta}_i \middle| \bm\theta \right)
    :=
    p\left(Y_i(\omega) = X_i(\omega; \hat{\bm\theta})
        \middle| \bm\theta \right)
    %\\
    \propto \exp \Bigg[
        - \frac{1}{2} \int_0^\infty \Bigg|
            \frac{\hat{\rho}_i}{\sigma_i(\hat{\bm\theta}_\mathrm{in})} e^{i (\hat\gamma_i - \omega \hat\tau_i)} \frac{H(\omega; \hat{\bm\theta}_\mathrm{in})}{S_i(\omega)}
    %\\
            - \frac{\rho_i}{\sigma_i(\bm\theta_\mathrm{in})} e^{i (\gamma_i - \omega \tau_i)} \frac{H(\omega; \bm\theta_\mathrm{in})}{S_i(\omega)}
        \Bigg|^2 \, d\omega
    \Bigg].
\end{equation}
\end{widetext}
%
If we further assume that the intrinsic parameters are equal to their \ac{ML} estimates, $\bm\theta_\mathrm{in} = \hat{\bm\theta}_\mathrm{in}$, then this reduces to what we call the autocorrelation likelihood,
%
\begin{widetext}
\begin{equation}\label{eq:autocor-likelihood}
    p\left(\hat\rho_i, \hat\gamma_i, \hat\tau_i
        \middle| \rho_i, \gamma_i, \tau_i \right) \propto
    \exp \left[ - \frac{1}{2}{\hat\rho_i}^2 - \frac{1}{2}{\rho_i}^2
        + \hat\rho_i \rho_i \Re \left\{ e^{i \tilde{\gamma}_i} a_i^*(\tilde{\tau}_i)
        \right\}
    \right],
\end{equation}
\end{widetext}
%
with $\tilde{\gamma}_i = \hat\gamma_i - \gamma_i$, $\tilde{\tau}_i = \hat\tau_i - \tau_i$, and the template's autocorrelation function $a_i(t; \bm\theta_\mathrm{in})$ defined as
%
\begin{equation}\label{eq:autocorrelation-function}
    a_i(t; \bm\theta_\mathrm{in}) := \frac{1}{{{\sigma_i}^2(\bm\theta_\mathrm{in})}} \int_0^\infty \frac{\left| H(\omega; \hat{\bm\theta}_\mathrm{in})\right|^2}{S_i(\omega)} e^{i \omega t} \,d\omega.
\end{equation}
%
Some example autocorrelation functions and corresponding likelihoods are shown in Fig.~\ref{fig:autocorr-likelihood}. To assemble the joint likelihood for the whole network, we form the product of the autocorrelation likelihoods from the individual detectors:
%
\begin{multline}
    p\left(\left\{\hat\rho_i, \hat\gamma_i, \hat\tau_i\right\}_i
        \middle| \left\{\rho_i, \gamma_i, \tau_i\right\}_i \right) \\
    \propto
    \exp \left[ - \frac{1}{2} \sum_i {\hat\rho_i}^2 - \frac{1}{2} \sum_i {\rho_i}^2
        + \sum_i \hat\rho_i \rho_i \Re \left\{ e^{i \tilde{\gamma}_i} a^*(\tilde{\tau}_i)
        \right\}
    \right].
\end{multline}

@@

We measured sky localization areas for each event as follows. First, we ranked the \ac{HEALPix} pixels by descending posterior probability. Then, we computed the cumulative sum of the pixels in that order. Finally, we searched for the index of the pixel of which the cumulative sum was equal to a given value: for example, 0.9 if we are interested in the 90\% credible area. That pixel index times the area per pixel is the area of the smallest region of the specified credible level. This area can be thought of measuring the precision of the sky localization: it is a measure of the scale of the posterior distribution.

@@

A more open\nobreakdashes-ended question is how to account for spin precession. The simulations in Ref.~\cite{FirstTwoYears} and in this paper featured extremely modest spins of $\chi \leq 0.05$, consistent with the fastest known pulsars in binaries \citep{2003Natur.426..531B,DetectingBNSSystemsWithSpin}. The signals were detected using a template bank that lacked spins entirely. Ref.~\cite{FirstTwoYearsSpin} shows that using nonspinning \ac{BNS} templates for parameter estimation has negligible impact on sky localization. However, if one or both companions are spinning as fast as a millisecond pulsar, $\chi \sim 0.4$ \cite{FastestSpinningMillisecondPulsar}, or even near breakup, $\chi \sim 0.7$, then the orbital plane may precess; in this case, spins can no longer be neglected for detection \cite{DetectingBNSSystemsWithSpin} and may also be important for parameter estimation. Since spinning \ac{BNS} searches are still an active area of development, \ac{BAYESTAR}'s sky localization accuracy in this regime should be reexamined in the future.

@@

This procedure has a shortcut called the Schur complement (see, for example, \citealt{numerical-recipes-inversion-partition}). Consider a partitioned square matrix $M$ and its inverse:
%
\begin{equation}
    M = \left(
        \begin{array}{cc}
            A & B \\
            C & D
        \end{array}
    \right), \qquad
    M^{-1} = \left(
        \begin{array}{cc}
            \tilde{A} & \tilde{B} \\
            \tilde{C} & \tilde{D}
        \end{array}
    \right).
\end{equation}
%
If $A$ and $B$ are square matrices, then the upper-left block of the inverse can be written as
%
\begin{equation}
    \tilde{A}^{-1} = A - B D^{-1} C.
\end{equation}
%
If we partition the $\mathcal{I}_\mathrm{net}$ similarly, the $A$ block consists of the first two rows and columns and $D$ is the lower right block that describes all other parameters. Because the Fisher matrix is symmetric, the off\nobreakdashes-diagonal blocks satisfy $C = B\transpose$. Then the Schur complement
%
\begin{equation}\label{eq:marginal-fisher-matrix}
    \mathcal{I}_\mathrm{marg} = A - B D^{-1} B\transpose
\end{equation}
%
gives us the information matrix marginalized over all parameters but $\delta\theta$ and $\delta\phi$.

>>>>

A more natural way is to plug the signal model from Eq.~(\ref{eq:signal-model}) evaluated at the \ac{ML} parameter estimates into the single-detector likelihood in Eq.~(\ref{eq:gaussian-likelihood-spa}):
%
\begin{multline}
    p\left(\hat{\bm\theta}_i \middle| \bm\theta \right)
    :=
    p\left(Y_i(\omega) = X_i(\omega; \hat{\bm\theta})
        \middle| \bm\theta \right)
    \\
    \propto \exp \Bigg[
        - \frac{1}{2} \int_0^\infty \Bigg|
            \frac{\hat{\rho}_i}{\sigma_i(\hat{\bm\theta}_\mathrm{in})} e^{i (\hat\gamma_i - \omega \hat\tau_i)} \frac{H(\omega; \hat{\bm\theta}_\mathrm{in})}{S_i(\omega)}
    \\
            - \frac{\rho_i}{\sigma_i(\bm\theta_\mathrm{in})} e^{i (\gamma_i - \omega \tau_i)} \frac{H(\omega; \bm\theta_\mathrm{in})}{S_i(\omega)}
        \Bigg|^2 \, d\omega
    \Bigg].
\end{multline}
%
If we further assume that the intrinsic parameters are equal to their \ac{ML} estimates, $\bm\theta_\mathrm{in} = \hat{\bm\theta}_\mathrm{in}$, then this reduces to what we call the autocorrelation likelihood,
%
\begin{multline}\label{eq:autocor-likelihood}
    p\left(\hat\rho_i, \hat\gamma_i, \hat\tau_i
        \middle| \rho_i, \gamma_i, \tau_i \right)
    \\
    \propto
    \exp \left[ - \frac{1}{2}{\hat\rho_i}^2 - \frac{1}{2}{\rho_i}^2
        + \hat\rho_i \rho_i \Re \left\{ e^{i \tilde{\gamma}_i} a_i^*(\tilde{\tau}_i)
        \right\}
    \right],
\end{multline}
%
with $\tilde{\gamma}_i = \hat\gamma_i - \gamma_i$, $\tilde{\tau}_i = \hat\tau_i - \tau_i$, and the template's autocorrelation function $a_i(t; \bm\theta_\mathrm{in})$ defined as
%
\begin{equation}\label{eq:autocorrelation-function}
    a_i(t; \bm\theta_\mathrm{in}) := \frac{1}{{{\sigma_i}^2(\bm\theta_\mathrm{in})}} \int_0^\infty \frac{\left| H(\omega; \hat{\bm\theta}_\mathrm{in})\right|^2}{S_i(\omega)} e^{i \omega t} \,d\omega.
\end{equation}
%
Some example autocorrelation functions and corresponding likelihoods are shown in Fig.~\ref{fig:autocorr-likelihood}. To assemble the joint likelihood for the whole network, we form the product of the autocorrelation likelihoods from the individual detectors:
%
\begin{multline}
    p\left(\left\{\hat\rho_i, \hat\gamma_i, \hat\tau_i\right\}_i
        \middle| \left\{\rho_i, \gamma_i, \tau_i\right\}_i \right) \\
    \propto
    \exp \left[ - \frac{1}{2} \sum_i {\hat\rho_i}^2 - \frac{1}{2} \sum_i {\rho_i}^2
        + \sum_i \hat\rho_i \rho_i \Re \left\{ e^{i \tilde{\gamma}_i} a^*(\tilde{\tau}_i)
        \right\}
    \right].
\end{multline}

@@

We measured sky localization areas for each event as follows. First, we ranked the \ac{HEALPix} pixels by descending posterior probability. Then, we computed the cumulative sum of the pixels in that order. Finally, we searched for the index of the pixel of which the cumulative sum was equal to a given value: for example, 0.9 if we are interested in the 90\% credible area. That pixel index times the area per pixel is the area of the smallest region of the specified credible level. This area can be thought of as measuring the precision of the sky localization: it is a measure of the scale of the posterior distribution.

@@

A more open\nobreakdashes-ended question is how to account for spin precession. The simulations in Ref.~\cite{FirstTwoYears} and in this paper featured extremely modest spins of $\chi \leq 0.05$, consistent with the fastest known pulsars in binaries \citep{2003Natur.426..531B,DetectingBNSSystemsWithSpin}. The signals were detected using a template bank that lacked spins entirely. Reference~\cite{FirstTwoYearsSpin} shows that using nonspinning \ac{BNS} templates for parameter estimation has negligible impact on sky localization. However, if one or both companions are spinning as fast as a millisecond pulsar, $\chi \sim 0.4$ \cite{FastestSpinningMillisecondPulsar}, or even near breakup, $\chi \sim 0.7$, then the orbital plane may precess; in this case, spins can no longer be neglected for detection \cite{DetectingBNSSystemsWithSpin} and may also be important for parameter estimation. Since spinning \ac{BNS} searches are still an active area of development, \ac{BAYESTAR}'s sky localization accuracy in this regime should be reexamined in the future.

@@

This procedure has a shortcut called the Schur complement (see, for example, \citeauthor{numerical-recipes-inversion-partition}~\cite{numerical-recipes-inversion-partition}). Consider a partitioned square matrix $M$ and its inverse:
%
\begin{equation}
    M = \left(
        \begin{array}{cc}
            A & B \\
            C & D
        \end{array}
    \right), \qquad
    M^{-1} = \left(
        \begin{array}{cc}
            \tilde{A} & \tilde{B} \\
            \tilde{C} & \tilde{D}
        \end{array}
    \right).
\end{equation}
%
If $A$ and $B$ are square matrices, then the upper-left block of the inverse can be written as
%
\begin{equation}
    \tilde{A}^{-1} = A - B D^{-1} C.
\end{equation}
%
If we partition the $\mathcal{I}_\mathrm{net}$ similarly, the $A$ block consists of the first two rows and columns and $D$ is the lower right block that describes all other parameters. Because the Fisher matrix is symmetric, the off\nobreakdashes-diagonal blocks satisfy $C = B\transpose$. Then the Schur complement
%
\begin{equation}\label{eq:marginal-fisher-matrix}
    \mathcal{I}_\mathrm{marg} = A - B D^{-1} B\transpose
\end{equation}
%
gives us the information matrix marginalized over all parameters but $\delta\theta$ and $\delta\phi$.
